---
title: "Music and mood"
author: "Akriti Acharya"
date: "2026-09-8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(dplyr)
library(readr)

```


# Intro: 

For my project, I’m using the mxmh survey dataset, which has 736 responses and 
33 columns. The data includes things like age, favorite genre, hours spent listening to music each day, and whether people listen while working. It also 
has mental health information such as anxiety, depression, insomnia, and OCD scores, plus whether people feel music improves, worsens, or has no effect on their mood.

This dataset is useful because it connects music habits with mental health and focus. That means I can look at questions like: do certain genres (like lo‑fi,
pop, or rap) relate to lower anxiety? Does listening while working help people focus? Or does the tempo (BPM) of music make a difference in how people feel?

The goal of my exploratory data analysis (EDA) is to get familiar with the
dataset, check patterns, and see what stands out. By looking at summaries, 
counts, and simple graphs, I’ll be able to understand the data better and
decide which research question makes the most sense to focus on.

# Looking at the data

```{r}
mxmh_survey_results <- read_csv("mxmh_survey_results.csv")
```
```{r}
# Look at the first few rows

head(mxmh_survey_results)

# Column names
names(mxmh_survey_results)

# Summary statistics
summary(mxmh_survey_results)

# Genre counts
table(mxmh_survey_results$`Fav genre`)
```

#Check for any missing values
```{r}
colSums(is.na(mxmh_survey_results))
```
#Getting rid of rows with missing ages
```{r}
mxmh_survey_results <- mxmh_survey_results%>%
  filter(!is.na(Age))
```

