---
title: "Music and mood"
author: "Akriti Acharya"
date: "2026-09-8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(dplyr)
library(readr)

```


# Intro: 

For my project, I’m using the mxmh survey dataset, which has 736 responses and 
33 columns. The data includes things like age, favorite genre, hours spent listening to music each day, and whether people listen while working. It also 
has mental health information such as anxiety, depression, insomnia, and OCD scores, plus whether people feel music improves, worsens, or has no effect on their mood.

This dataset is useful because it connects music habits with mental health and focus. That means I can look at questions like: do certain genres (like lo‑fi,
pop, or rap) relate to lower anxiety? Does listening while working help people focus? Or does the tempo (BPM) of music make a difference in how people feel?


# Looking at the data

```{r}
mxmh_survey_results <- read_csv("mxmh_survey_results.csv")
```
```{r}
# Looking at the first few rows

head(mxmh_survey_results)

# Column names
names(mxmh_survey_results)

# Summary statistics
summary(mxmh_survey_results)

# Genre counts
table(mxmh_survey_results$`Fav genre`)
```

#Check for any missing values
```{r}
colSums(is.na(mxmh_survey_results))
```
#Summary stats for quantative variable
```{r}
summary(mxmh_survey_results[, c("Age", "Hours per day", "BPM", 
                                "Anxiety", "Depression", "Insomnia", "OCD")])
```


#Cleaning outliers
```{r}
mxmh_survey_results <- mxmh_survey_results %>%
  filter(BPM < 300 | is.na(BPM))
```


# Distribution of Favorite Genres

```{r}
library(ggplot2)

ggplot(mxmh_survey_results, aes(x = `Fav genre`)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Counts of Favorite Genres",
       x = "Genre",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Average Anxiety Score by Genre

```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(mean_anxiety = mean(Anxiety, na.rm = TRUE)) %>%
  ggplot(aes(x = `Fav genre`, y = mean_anxiety)) +
  geom_col(fill = "purple") +
  labs(title = "Average Anxiety Score by Genre",
       x = "Genre",
       y = "Mean Anxiety") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

