---
title: "Music and mood"
author: "Akriti Acharya"
date: "2026-09-8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)

```


# Introduction: 

For my project, I’m using the mxmh survey dataset, which has 736 responses and 33 columns. The data includes things like age, favorite genre, hours spent listening to music each day, and whether people listen while working. It also 
has mental health information such as anxiety, depression, insomnia, and OCD scores, plus whether people feel music improves, worsens, or has no effect on their mood.

This dataset is useful because it connects music habits with mental health and focus. That means I can look at questions like: do certain genres (like lo‑fi,pop, or rap) relate to lower anxiety? Does listening while working help people focus? Or does the tempo (BPM) of music make a difference in how people feel?

#Background

Music and Emotional Regulation

Music engages brain reward pathways, releasing dopamine and reducing stress and anxiety. As the Healthline article explains, listening to preferred music can boost motivation and pleasure, though certain genres may reinforce depressive moods when listeners are already vulnerable. This dual effect highlights the complexity of music’s role in mental health.

Cognition, Identity, and Clinical Applications

The Home Monitor on Psychology article emphasizes music’s role in cognition and identity. Research shows that self‑selected music strongly activates reward circuits tied to personal values, explaining its deep emotional impact. Clinical applications include neurologic music therapy, which has been used to restore speech and movement in patients with brain injuries. Studies also show music can regulate memory, imagination, and cultural identity, making it a powerful tool for psychological health.


# Looking at the data

```{r}
mxmh_survey_results <- read_csv("mxmh_survey_results.csv")
```
```{r}
# Looking at the first few rows

head(mxmh_survey_results)

# Column names
names(mxmh_survey_results)

# Summary statistics
summary(mxmh_survey_results)

# Genre counts
table(mxmh_survey_results$`Fav genre`)

ggplot(mxmh_survey_results,aes(x=BPM))+ 
  geom_boxplot()

```

```{r}
mxmh_survey_results %>%
  filter(BPM>1000)

is.na(mxmh_survey_results$BPM)<- mxmh_survey_results$BPM > 1000
```
```{r}
ggplot(mxmh_survey_results,aes(x=BPM))+ 
  geom_boxplot()
```

#Check for any missing values
```{r}
colSums(is.na(mxmh_survey_results))
```
#Summary stats for quantative variable
```{r}
summary(mxmh_survey_results[, c("Age", "Hours per day", "BPM", 
                                "Anxiety", "Depression", "Insomnia", "OCD")])
```

# Distribution of Favorite Genres

```{r}
library(ggplot2)

ggplot(mxmh_survey_results, aes(x = `Fav genre`)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Counts of Favorite Genres",
       x = "Genre",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
We can see that rock is the most common favorite genre, followed by Pop and Metal, while Gospel and Latin have the fewest responses.This shows which genres are most popular in the dataset.


# Average Anxiety Score by Genre

```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(mean_anxiety = mean(Anxiety, na.rm = TRUE)) %>%
  ggplot(aes(x = `Fav genre`, y = mean_anxiety)) +
  geom_col(fill = "purple") +
  labs(title = "Average Anxiety Score by Genre",
       x = "Genre",
       y = "Mean Anxiety") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

 Folk listeners reported the highest average anxiety scores, while Latin listeners reported the lowest. This suggests that certain genres may be linked with higher or lower levels of anxiety.

# Depression by Genre 
```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(mean_depression = mean(Depression, na.rm = TRUE)) %>%
  ggplot(aes(x = `Fav genre`, y = mean_depression)) +
  geom_col(fill = "darkred") +
  labs(title = "Average Depression Score by Genre",
       x = "Genre",
       y = "Mean Depression") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
This chart shows that lo‑fi, hip hop, and EDM(electronic dance music) listeners report higher average depression scores, while gospel and Latin music  listeners report the lowest. The chart suggests certain genres may be linked with higher or lower depression levels.



# Using median instead of mean for the visuals 

Anxiety by genre (using meadian)
```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(median_anxiety = median(Anxiety, na.rm = TRUE)) %>%
  ggplot(aes(x = `Fav genre`, y = median_anxiety)) +
  geom_col(fill = "purple") +
  labs(title = "Median Anxiety Score by Genre",
       x = "Genre",
       y = "Median Anxiety") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Using the median instead of the mean gives a clearer picture of the typical listener’s anxiety level, since it reduces the influence of extreme outliers. Compared to the mean chart, small differences appear: genres like Rock and Lofi show higher median anxiety than their mean suggested, while Rap shows lower. This indicates that outliers were affecting the averages, and the median provides a more reliable summary of how anxiety varies across genres.



#Depression by genre (using median)
```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(median_depression = median(Depression, na.rm = TRUE)) %>%
  ggplot(aes(x = `Fav genre`, y = median_depression)) +
  geom_col(fill = "darkred") +
  labs(title = "Median Depression Score by Genre",
       x = "Genre",
       y = "Median Depression") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Switching from mean to median highlights some differences in how genres compare. Lofi listeners show the highest median depression scores, even higher than their mean suggested, while Gospel and Latin remain the lowest. Hip hop stays high, but the median chart makes Lofi stand out more clearly as the top. This shows that extreme values were influencing the averages, and the median provides a more representative summary of typical scores across genres.



#Distributions of Quantative Variables


#Age distribution

```{r}
ggplot(mxmh_survey_results, aes(x = Age)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  labs(title = "Age Distribution", x = "Age", y = "Count")

ggplot(mxmh_survey_results, aes(y = Age)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Age Boxplot", y = "Age")
```

The age distribution is right‑skewed, with most participants between 20–30 years old and a clear peak near 25. Counts steadily decrease as age increases, with very few respondents above 60.

The boxplot confirms this skew, showing the interquartile range centered in the mid‑20s and numerous outliers extending into older ages. This suggests the dataset is dominated by younger participants, while older ages appear as outliers that pull the mean upward compared to the median.



#Hours per Day Listenting to music 
```{r}
ggplot(mxmh_survey_results, aes(x = `Hours per day`)) +
  geom_histogram(binwidth = 0.5, fill = "purple") +
  labs(title = "Hours per Day Distribution", x = "Hours per Day", y = "Count")

ggplot(mxmh_survey_results, aes(y = `Hours per day`)) +
  geom_boxplot(fill = "purple") +
  labs(title = "Hours per Day Boxplot", y = "Hours per Day")

```
The distribution is right‑skewed, with most participants listening between 1–5 hours daily and a clear peak around 2–3 hours. 
The boxplot shows the interquartile range centered in this range, with the median near 3 hours. Several extreme outliers appear above 10 hours, extending up to 25, which pull the mean upward compared to the median. This indicates that while typical listening behavior is moderate, a small group of heavy listeners significantly increases the spread.



#Anxiety Distribution
```{r}
ggplot(mxmh_survey_results, aes(x = Anxiety)) +
  geom_histogram(binwidth = 1, fill = "orange") +
  labs(title = "Anxiety Score Distribution", x = "Anxiety Score", y = "Count")

ggplot(mxmh_survey_results, aes(y = Anxiety)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Anxiety Score Boxplot", y = "Anxiety Score")


```
Distribution: The anxiety distribution is left‑skewed, with most participants reporting scores between 6 and 9. The tallest bars appear around 7–8, showing that higher anxiety scores are more common in this dataset. Lower scores (0–3) occur much less frequently, creating a long tail on the left side. This pattern suggests that the sample overall tends toward moderate to high anxiety levels.

Box plot:
The boxplot shows the median anxiety score around 6, with the middle 50% of scores (IQR) ranging roughly from 4 to 8. The whiskers extend from 0 to 10, indicating the full range of responses. There are no extreme outliers, but the wide spread shows substantial variability in anxiety levels across participants. The higher concentration of values in the upper quartile aligns with the histogram’s peak around 7–8. 



#Depression Distribution
```{r}
ggplot(mxmh_survey_results, aes(x = Depression)) +
  geom_histogram(binwidth = 1, fill = "darkred") +
  labs(title = "Depression Score Distribution", x = "Depression Score", y = "Count")

ggplot(mxmh_survey_results, aes(y = Depression)) +
  geom_boxplot(fill = "darkred") +
  labs(title = "Depression Score Boxplot", y = "Depression Score")
```
Distribution: The depression distribution is multi‑modal, meaning it has several peaks rather than one clear dominant range. You can see noticeable clusters around 0–2 and again around 7–8, showing that participants tend to fall into either lower or higher depression score groups. Middle scores (3–5) appear less frequent, creating a dip in the center of the distribution. This pattern suggests the dataset contains two distinct groups: one with low depression symptoms and another with higher levels.

Box Plot: The boxplot shows the median depression score around 5, with the interquartile range stretching from roughly 2.5 to 7.5. This wide IQR indicates substantial variability in depression levels across the sample. The whiskers extend across the full range (0–10), and there are no extreme outliers, meaning the variation is part of the natural spread of responses. The boxplot’s wide middle section aligns with the histogram’s two clusters, showing that depression scores are not centered around a single typical value.




#Insomnia Distribution
```{r}

ggplot(mxmh_survey_results, aes(x = Insomnia)) +
  geom_histogram(binwidth = 1, fill = "skyblue") +
  labs(title = "Insomnia Score Distribution", x = "Insomnia Score", y = "Count")

ggplot(mxmh_survey_results, aes(y = Insomnia)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Insomnia Score Boxplot", y = "Insomnia Score")
```
Distribution: The insomnia distribution is strongly right‑skewed, with the highest concentration of scores at 0. The count drops steadily as insomnia scores increase, showing that most participants report little to no insomnia symptoms. Higher scores (7–10) appear much less frequently, forming a long tail on the right side. This pattern suggests that insomnia is not widespread in the dataset, but a smaller group of participants experiences more severe symptoms.

Box plot:  The boxplot shows the median insomnia score around 3, with the interquartile range stretching from roughly 1 to 6. This indicates moderate variability among participants. The whiskers extend across the full range (0–10), showing that while most scores cluster toward the lower end, the dataset includes individuals with severe insomnia. The combination of a low median and a long right tail aligns with the histogram’s strong right skew.



#OCD Distribution
```{r}
ggplot(mxmh_survey_results, aes(x = OCD)) +
  geom_histogram(binwidth = 1, fill = "gold") +
  labs(title = "OCD Score Distribution", x = "OCD Score", y = "Count")

ggplot(mxmh_survey_results, aes(y = OCD)) +
  geom_boxplot(fill = "gold") +
  labs(title = "OCD Score Boxplot", y = "OCD Score")


```
Distribution:  The OCD distribution is strongly right‑skewed, with the highest concentration of scores at 0. The count drops sharply as scores increase, showing that most participants report very low OCD symptoms. Higher scores (7–10) occurless frequently, forming a long right tail. This pattern indicates that OCD symptoms are uncommon in the dataset, but a smaller group of participants experiences more severe levels.

Box plot:  The boxplot shows the median OCD score around 2, with the interquartile range stretching from roughly 0 to 5. This suggests that half of the participants fall within the low‑to‑moderate range. The whiskers extend across the full range, showing that while most scores cluster toward the lower end, the dataset includes individuals with high OCD scores. The combination of a low median and a long right tail aligns with the histogram’s strong right skew.




#Music effects on mood bar chart
```{r}
ggplot(mxmh_survey_results, aes(x = `Music effects`)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Music Effects on Mood",
       x = "Effect",
       y = "Count")
```
Looking at the chart, most participants said music improves their mood, with far fewer reporting no effect or worsened mood. This indicates that overall, music is strongly associated with positive emotional wellbeing.

```{r}
ggplot(mxmh_survey_results, aes(x = `While working`)) +
  geom_bar(fill = "darkblue") +
  labs(title = "Listening to Music While Working",
       x = "Response",
       y = "Count")
```
Most participants reported listening to music while working, with the “Yes” category being the largest group. A smaller portion answered No, and very few responses were missing (NA). This shows a clear preference for using music during tasks, suggesting that music is commonly integrated into work or study routines.


#Anxiety vs Depression by Genre (using median)
```{r}
mxmh_survey_results %>%
  group_by(`Fav genre`) %>%
  summarise(
    median_anxiety = median(Anxiety, na.rm = TRUE),
    median_depression = median(Depression, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(median_anxiety, median_depression),
               names_to = "Measure",
               values_to = "Score") %>%
  ggplot(aes(x = `Fav genre`, y = Score, fill = Measure)) +
  geom_col(position ="dodge" ) +
  labs(title = "Median Anxiety vs. Depression Score by Genre",
       x = "Genre",
       y = "Median Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This chart compares median anxiety and depression scores side by side for each genre. It shows that some genres, like Lofi and Hip hop, have high median scores for both measures, while Gospel and Latin remain consistently low. Looking at both together highlights which genres are linked with higher or lower levels of mental health concerns, and using the median ensures the results are more representative of typical listeners rather than being skewed by outliers.



